P3064R2: How to Avoid OOTA Without Really Trying

### 2024-09-01 paper/ootavacuous.tex
paper/ootavacuous: Wordsmith Section 1 ("Introduction and Background")
Drop the section introducing herd7 and adjust the examples
accordindly.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} having final values of 42, despite the fact
that there is nothing in the initial values or the executable code to
support such an outcome.
First, P0's line~1 loads from \co{x} into \co{r1}, claiming to read the value
of P1's line-2 store rather than \co{x}'s initial value and somehow
obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles,
the C++ standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~6 to line~7 and
another from line~12 to line~13.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	See Section~\ref{sec:Hardware Architecture and Design}
	for more discussion about hardware speculation.}
the stores in lines~7 and~13 cannot take place until the CPU
knows what values are loaded by lines~6 and~12, respectively.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder
lines~12 and~13, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior.

@@

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
early 2024, Mark Batty is persisting with modular relaxed dependencies.

@@

\subsection{Code-Analysis Tool}
\label{sec:Code-Analysis Tool}

This paper will use the \co{herd7}\footnote{
	Available at \path{https://github.com/herd/herdtools7}.}
tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments.
In some cases its output will include executions with OOTA cycles,
on occasion reporting undefined values for the variables involved in
the cycle.\footnote{
	This happens only some of the time because of idiosyncrasies
	in the algorithm used by \co{herd7}'s self-consistency solver.}

\begin{listing}[tbp]
@@ DisplayLitmus litmus/oota-ctrl.litmus @@
\caption{OOTA Cycle}
\label{lst:OOTA Cycle}
\end{listing}

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that under loose C++ has an OOTA cycle (although
the cycle is of course prohibited in full C++).
This section describes the fragment, thereby giving an overview of
the \co{herd7} tool.

The first line identifies it as a C-language litmus test and gives it
a name, and this name identifies the litmus test's source file within
the \path{litmus} directory in the \url{https://github.com/paulmckrcu/oota}
repository.
Lines~2--5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
(Variables that are not explicitly initialized are initialized
to zero by default.)
Lines~7--13 define the first thread, \co{P0()}, and lines~15--21
define the second thread, \co{P1()}.\footnote{
	The ``P'' stands for ``process'', which is \co{herd7}'s name
	for ``thread''.}
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each thread may access, in this case, \co{x}
and \co{y}.
The body of each thread contains C++ code,
written in a slightly stilted manner to keep the load operations separate
from the rest and because \co{herd7}'s knowledge of C++ is limited.

Line~23 has a \co{locations} clause, which causes
\co{herd7} to dump out the final values of \co{x} and \co{y}.
Finally, line~24 specifies an \co{exists} clause, which gives a
condition to check for the final values of the specified variables.
The \co{0:} prefix denotes a variable local to \co{P0()} and
the \co{1:} prefix denotes a variable local to \co{P1()}.
The \co{/\\} is a boolean AND, and the \co{=} signs are equality comparisons.
If a variable appears in the \co{exists} clause then the final value of
that variable constitutes observable behavior.

\begin{listing}[tbp]
@@ RunLitmus litmus/oota-ctrl.litmus @@
\caption{OOTA Cycle, \co{herd7} Output}
\label{lst:OOTA Cycle, herd7 Output}
\end{listing}

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both threads load the value 42.
Normally these lines would include only those variables mentioned
in the \co{exists} clause, but because of the \co{locations} clause
the values of \co{x} and \co{y} are also listed, which
can be helpful for debugging.
% @@@ As we will see, this outcome is not possible in real-world
% @@@ implementations due to physical constraints of which \co{herd7}
% @@@ is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
\caption{Simple Reordering as Litmus Test}
\label{lst:Simple Reordering as Litmus Test}
\end{listing}

Later examples will usually combine the litmus test and the \co{herd7}
output into one listing, as shown in
Listing~\ref{lst:Simple Reordering as Litmus Test},
which recasts
Listing~\ref{lst:Simple Reordering}
into litmus-test form.
Listing~\ref{lst:Simple OOTA}
can also be recast as a \co{herd7} litmus test, as shown in
Listing~\ref{lst:Causality Test Case 4}
on page~\pageref{lst:Causality Test Case 4}.
Other OOTA-related litmus tests may be found in
Appendix~\ref{app:Illustrative Litmus Tests}.

>>>>

Listing~\ref{lst:Simple OOTA} shows a simple example where an OOTA
cycle might result in all of \co{x}, \co{y}, \co{r1}, and \co{r2}
having final values of 42, despite the fact that there is nothing in
the default-zero initial values or the executable code to support such
an outcome~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
The ``=\textsubscript{rlx}'' denotes a relaxed access, with variables
whose names are ``\co{r}'' followed by a number denoting registers,
and other names denoting shared variables.
First, P0's line~1 does a relaxed atomic load from \co{x} into \co{r1},
reading the value of P1's line-2 store rather than \co{x}'s initial
value of zero and somehow obtaining 42.
Second, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Third, P1's line~1 loads 42 from \co{y} to \co{r2}.
Fourth and finally, P1's line~2 stores 42 to \co{x}, justifying the value
loaded by P0's line~1.

Because nothing else in the C++ memory model rules out such OOTA cycles
(as indicated by the ``sometimes satisfied'' in the figure), the C++
standard explicitly prohibits them in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:
\begin{quote}
	Implementations should ensure that no “out-of-thin-air” values
	are computed that circularly depend on their own computation.
\end{quote}
The standard's prohibition of OOTA is of course important, for example, to
prevent misapplication of the as-if rule in oracular C++ implementations,
but those of us writing code in the real world must rely on actual C++
implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data-dependency ordering in
weakly ordered systems.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations require checks on such
	speculation, and these checks must be based on actual values
	loaded.}

In Listing~\ref{lst:Simple OOTA}
there is a \emph{semantic dependency} from line~1 to line~2 in both
P0 and P1.
(Roughly speaking, there is a semantic dependency from a given load
to a given store when \emph{all other things being equal, a change in the
value loaded can change the value stored or prevent the store from
occurring at all.}
Here the dependencies are trivial because the values stored simply
\emph{are} the values that were loaded.)
Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	See Section~\ref{sec:Hardware Architecture and Design}
	for more discussion about hardware speculation.}
the stores on line~2 cannot take place until the P0's and P1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder P1's
lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior,
as indicated by the ``sometimes satisfied'' in the figure.

\subsubsection{Non-Trivial OOTA}
\label{sec:Non-Trivial OOTA}

@@

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-ctrl.litmus @@
\caption{OOTA Cycle}
\label{lst:OOTA Cycle}
\end{listing}

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that under loose C++ has an OOTA cycle, though
the cycle is of course prohibited in full C++.
This example differs from that of Section~\ref{sec:Simple OOTA Cycle}
in that line~3 of both processes stores the value 42, but only if
line~2 of a given process determines that the value loaded was 42,
keeping in mind that variables are default-initialized to zero.

This OOTA cycle is constructed in the following manner.
First, P0's and P1's line~1 load from \co{y} and \co{x}, respectively,
both somehow returning the value 42.
Second, P0's and P1's line~2 determine that the values loaded were both
42, so each process advances to its line~3.
Third, P0's and P1's line~3 stores the value 42 that was (somehow)
clairvoyantly loaded by the other process's line~1.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
2024, Mark Batty is persisting with modular relaxed dependencies.
@@@ cite web site.

@@
